---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bnmonitor

`bnmonitor` is a package for sensitivity analysis and robustness in Bayesian networks (BNs).

## Installation 

The package `bnmonitor` can be installed from GitHub using the command 
```{r warning = FALSE, message = FALSE}
# install.packages("devtools")
devtools::install_github("manueleleonelli/bnmonitor")
```
and loaded in R with
```{r}
library(bnmonitor)
```
Note that `bnmonitor` requires the package `gRain` which, while on CRAN, depends on packages that are on Bioconductor both directly and through the `gRbase` package, which depends on `RBGL`:
```{r eval =F}
install.packages("BiocManager")
BiocManager::install(c("graph", "Rgraphviz", "RBGL"))
install.packages("gRain")
```

## Overview

`bnmonitor` provides a suite of function to investigate either a data-learnt or an expert elicited BN. Its functions can be classified into three main areas:

 - *Robustness in discrete BNs*: checking how well a BN represents a dataset;
 - *Sensitivity in discrete BNs*: assessing the effect of changes in the discrete BN's probabilities;
 - *Sensitivity in continuous BNs*: assessing the effect of changes in the continuous BN's probabilities, either in the standard or model-preserving framework

<!-- The prequential diagnostics examine the forecasts that flow from a model in sequence. -->
<!-- Each monitor given below indicates the probability of a particular observation based on the previous observations and the model structure.  -->
<!-- In the prequential mindset, we compute a probability of each subsequent observation based on all previous data points.  -->
<!-- These observations are then scored, and in this package we use the logarithmic score function. -->
<!-- The observations are then standardized to give a z-score statistic.  -->
<!-- Following the recommendation of Cowell (2007), scores indicate a poor fit where |z| > 1.96  -->

<!-- We demonstrate the efficacy of the prequential monitors with the Asia data set from the bnlearn package. Details of the variables (nodes) can be found in the documentation for bnlearn. -->

## Robustness

Consider the `asia` dataset available from `bnlearn`. Details of the variables can be found in the `bnlearn` documentation.

``` {r}
library(bnlearn)
data(asia)
summary(asia)
```

Consider two candidate BN models (Lauritzen and Spiegelhalter, 1988), which only differ in the edge from `S` to `D`.
``` {r out.width="50%"}
asia_bn <- bnlearn::model2network("[A][S][T|A][L|S][B|S][D|B:E][E|T:L][X|E]")
bnlearn::graphviz.plot(asia_bn) 
asia_bn_alt <-  bnlearn::model2network("[A][S][T|A][L|S][B|S][E|T:L][X|E][D|B:E:S]")
bnlearn::graphviz.plot(asia_bn_alt) 
```

### Global monitor

<!-- The global monitor is equivalent to the Bayes Factor and offers a good assessment of the model as a whole.  -->
<!-- It is primarily useful for differentiating between candidate models as the global monitor pinpoints the nodes with different contributions for two candidate models. -->
<!-- We can assess the log likelihood contribution from each of the nodes and determine which nodes contribute the most to the global monitor. -->
<!-- global.monitor.graph provides a quick visual for the entire network.  -->
<!-- The darker the color, the more substantial the contribution of that node to the log likelihood. -->

<!-- The global monitor assesses the probability of observing the data given the model and as such does not particularly fit the prequential view.  -->
<!-- However, it is a quick and useful first diagnostic assessment. -->

A first useful diagnostic is the `global_monitor`, reporting the contribution of each vertex to the log-likelihood of the model.

``` {r}
glob_asia <- node_monitor(dag = asia_bn, df = asia, alpha = 3)
glob_asia_alt <- node_monitor(dag = asia_bn_alt, df = asia, alpha = 3)
glob_asia
glob_asia_alt
```

In the alternative model, Dysnopea contributes more to the log-likelihood. 

Global monitors can be plotted giving a quick view of the decomposition of the log-likelihood. The darker the color, the more substantial the contribution of a vertex.

```{r out.width = "50%"}
plot(glob_asia)
```

### Node monitor 

There are two variants of node monitors.

  - The marginal node monitor computes the probability of the $i$th observation in the data set in turn after passing the evidence of the $i-1$th cases in the data set.
  
  - The conditional node monitor computes the probability of the $i$th observation in the data set after passing evidence of the $i-1$th cases in the data set, and the evidence for all nodes in the $i$th case except the node of interest.

As a quick survey of the nodes, the `node_monitor` command computes the marginal and conditional monitors for the final observation in the data set. 

``` {r echo=TRUE, message = FALSE, warning = FALSE}
node_asia <- final_node_monitor(dag = asia_bn, df = asia)
node_asia
```

The scores indicate a poor fit of the probability distributions specified for the Smoking, Bronchitis, and Dysnopea nodes, since these are larger than 1.96 in absolute value. Plots can also be created to give a visual counterpart of the node monitors.

```{r out.width="50%"}
plot(node_asia, which = "marginal")
```

<!-- The modeller may decide which of these distributions is of the most interest.  -->
<!-- For the purposes of this vignette, we assume that we would like to check the forecasts for Dysnopea.  -->

As an illustration we investigate further the fit of the variable Dysnopea.

The sequential marginal monitor `seq_marg_monitor` gives us a closer look at which particular forecasts in the data set might cause this poor fit. We examine the sequential monitor for both candidate models.


``` {r warning=FALSE}
seq_asia <- seq_marg_monitor(dag = asia_bn, df = asia, node.name = "D")
seq_asia_alt <- seq_marg_monitor(dag = asia_bn_alt, df = asia, node.name = "D")
seq_asia
seq_asia_alt
```


```{r out.width="50%",warning=FALSE}
library(gridExtra)
grid.arrange(plot(seq_asia),plot(seq_asia_alt),ncol=2)
```

Both monitors indicate that for some observations there is a poor fit (score above 1.96 in absolute value). In particular for the alternative models the marginal monitor has larger values in absolute value. 

A similar analysis can be conducted with `seq_marg_monitor`, which would show that the model fits well (not reported here).


### Parent Child monitor 

Once a vertex has been identified as a poor fit, further investigation can be carried out to check for which values of the parents the model provides a bad representation. This can be achieved with the `seq_pa_ch_monitor` function.

As an illustration consider the `asia_bn` BN, the vertex `D` (Dysnopea), the parent variable `B` (Bronchitis) which can take values `yes` and `no`.

``` {r warning=FALSE, out.width="50%"}
asia_pa_ch1 <- seq_pa_ch_monitor(dag = asia_bn, df = asia, node.name = "D", pa.names =  "B", pa.val =  "yes", alpha = 3)
asia_pa_ch1
asia_pa_ch2 <- seq_pa_ch_monitor(dag = asia_bn, df = asia, node.name = "D", pa.names = "B", pa.val = "no", alpha = 3)
asia_pa_ch2
grid.arrange(plot(asia_pa_ch1),plot(asia_pa_ch2), ncol = 2)
```

For this model, Dysnopea is adequately modeled for both values of Bronchitis, since most scores largely fall in the recommended range.

### Influential observations

The last robustness tool is the absolute value of the log-likelihood ratio between a model learnt without one observation and the one learnt with the full dataset. Larger values are associated to atomic events which influence the structural learning. 


```{r warning=FALSE, out.width="50%"}
influence <- influential_obs(dag = asia_bn, df = asia, alpha = 3)
head(influence)
plot(influence)
```

## Sensitivity in Discrete Bayesian Networks

The functionalities of `bnmonitor` for sensitivity analysis in Discrete Bayesian Networks are illustrated using the `fire_alarm` network bundled within the package representing a simple alarm system for fire accidents. 

```{r out.width="50%"}
graphviz.plot(fire_alarm)
```

The network consists of 6 nodes (`Fire`, `Smoke`, `Tampering`, `Alarm`, `Leaving` and `Report`) of either two or three levels. 

### Covariation schemes

The aim of a sensitivity analysis is to assess how parameter changes affect output probabilities of interest. When one parameter changes than others need to be adjusted and this can be done in many ways, including  `proportional`, `uniform` and `order-preserving` schemes. Suppose we vary the probability that `Alarm = FALSE` given that `Fire = TRUE` and `Tampering = FALSE` from 0.01 to 0.1. The resulting probabilities from the proportional covariation scheme can be computed as follows.

```{r}
proportional_covar(bnfit = fire_alarm, node = "Alarm", value_node = "FALSE", value_parents = c("TRUE","FALSE"), new_value = 0.1)$Alarm$prob
```

The functions `uniform_covar` and `orderp_covar` have the same inputs and return the BN where probabilities are updated using uniform and order-preserving covariation, respectively.

### Sensitivity functions

Suppose we are interested in the probability that `Report = TRUE` and how this varies with respect to the probability that `Alarm = FALSE` conditional on `Fire = TRUE` and `Tampering = FALSE`. This can be computed with the function `sensitivity`.

```{r out.width="50%", warning = FALSE}
sens <- sensitivity(bnfit = fire_alarm, interest_node = "Report", interest_node_value =  "TRUE" , node = "Alarm" , value_node = "FALSE", value_parents = c("TRUE","FALSE"), new_value = "all", covariation = "all")
plot(sens)
```

For all covariation schemes as the conditional probability of `Alarm = FALSE` increases, the probability of `Report = TRUE` decreases. The uniform covariation scheme is reported in red, the proportional in green and the order-preserving in blue.

The function sensitivity also allows for conditional probabilities of interest, not only marginals, which can be set by the options `evidence_nodes` and `evidence_states`.

For such changes in the probability distribution of a BN, we can compute the distance between the original and the varied networks using the Chan-Darwiche distance (`CD`) and the Kullback-Leibler divergence (`KL`). As an example consider `Alarm = MALFUNCTIOn` conditional on `Fire = TRUE` and `Tampering = FALSE`.

```{r out.width="50%", warning = FALSE}
cd_distance <- CD(fire_alarm, node = "Alarm", value_node = "MALFUNCTION", value_parents = c("TRUE","FALSE"), new_value = "all", covariation = "all")
kl_divergence <- KL(fire_alarm, node = "Alarm", value_node = "MALFUNCTION", value_parents = c("TRUE","FALSE"), new_value = "all", covariation = "all")
grid.arrange(plot(cd_distance), plot(kl_divergence),ncol=2)
```

As expected proportional covariation has the smallest distance/divergence of all covariation schemes.

### Sensitivity queries

Another task in sensitivity analysis is to identify parameter changes that meet a specific constraint. As an example suppose we want the probability `Report = TRUE` to be equal to 0.4. Which parameters could we change to obtain such a probability? The answer can be obtained using the function `sensquery`.

```{r}
sensquery(bnfit = fire_alarm, interest_node = "Report", interest_node_value = "TRUE", new_value = 0.4)
```

There are five parameter changes that would lead to a probability of `Report = TRUE` of 0.4. These are ordered by the value of the CD distance between the original and the new network, from smallest to largest.

## Gaussian Bayesian networks

The functionalities of `bnmonitor` for sensitivity analysis in Gaussian BNs are illustrated next using the `mathmarks` dataset bundled within the package. 
```{r}
data(mathmarks)
head(mathmarks)
```
The data includes the grades (out of 100) of students in five maths exams: mechanics, vectors, algebra, analysis and statistics.

The structure of a BN for this data is first learnt using the package `bnlearn` and the maximum likelihood estimate of its parameters is computed and stored in `bnfit`.
```{r out.width="50%"}
bn <- hc(mathmarks)
plot(bn)
bnfit <-bn.fit(bn, mathmarks)
```

To start the sensitivity analysis for the parameters of the learnt BN, one first need to transform `bnfit` to objects of class `GBN` (for standard sensitivity analysis) and `CI` (for model-preserving sensitivity). This can be done using the functions `bn2gbn` and `bn2ci` respectively.
```{r}
gbn <- bn2gbn(bnfit)
ci <-  bn2ci(bnfit)
c(class(gbn), class(ci))
```


### Perturbation of the mean vector

A varied Gaussian BN after a perturbation of an entry of the mean vector can be obtained with the function `mean_var`, which can only be applied to an object of class `GBN`. Below, we vary the fifth entry of the mean vector (statistics) by an additive factor 10.

```{r}
mean_varied <- cbind(gbn$order, round(gbn$mean, 2),round(mean_var(gbn = gbn, entry = 5, delta = 10)$mean, 2))
colnames(mean_varied) <- c("Course", "Original Mean", "Varied Mean")
mean_varied
```

The overall effect of such variations can be assessed in terms of dissimilarity measures: the Kullback-Leibler divergence (`KL`) and Jeffrey's divergence (`Jeffreys`). For instance, let's see what's the effect of variations in the mean of the statistics exam.

```{r out.width="50%", warning = FALSE}
kl_var5 <- KL(gbn, where = "mean", entry = 5, delta = seq(-10,10,0.1))
jef_var5 <- Jeffreys(gbn, where = "mean", entry = 5, delta = seq(-10,10,0.1))
grid.arrange(plot(kl_var5), plot(jef_var5), ncol = 2)
```

More interestingly, one can check the different effect of variations of different parameters (code not shown).

```{r echo = FALSE, out.width="50%", warning = FALSE}
library(ggplot2)
mean_var1 <- KL(gbn, "mean", entry=1, delta = seq(-10,10,0.1))$KL
mean_var2 <- KL(gbn, "mean", entry=2, delta = seq(-10,10,0.1))$KL
mean_var3 <- KL(gbn, "mean", entry=3, delta = seq(-10,10,0.1))$KL
mean_var4 <- KL(gbn, "mean", entry=4, delta = seq(-10,10,0.1))$KL
mean_var5 <- KL(gbn, "mean", entry=5, delta = seq(-10,10,0.1))$KL
out <- data.frame(delta = rep(seq(-10,10,0.1),5), KL = c(mean_var1[,2],mean_var2[,2],mean_var3[,2],mean_var4[,2],mean_var5[,2]), Entry = c(rep("mechanics",length(seq(-10,10,0.1))),rep("vectors",length(seq(-10,10,0.1))),rep("algebra",length(seq(-10,10,0.1))),rep("analysis",length(seq(-10,10,0.1))),rep("statistics",length(seq(-10,10,0.1)))))
mean_var1J <- Jeffreys(gbn, "mean", entry=1, delta = seq(-10,10,0.1))$Jeffreys
mean_var2J <- Jeffreys(gbn, "mean", entry=2, delta = seq(-10,10,0.1))$Jeffreys
mean_var3J <- Jeffreys(gbn, "mean", entry=3, delta = seq(-10,10,0.1))$Jeffreys
mean_var4J <- Jeffreys(gbn, "mean", entry=4, delta = seq(-10,10,0.1))$Jeffreys
mean_var5J <- Jeffreys(gbn, "mean", entry=5, delta = seq(-10,10,0.1))$Jeffreys
out1 <- data.frame(delta = rep(seq(-10,10,0.1),5), Jeffreys = c(mean_var1J[,2],mean_var2J[,2],mean_var3J[,2],mean_var4J[,2],mean_var5J[,2]), Entry = c(rep("mechanics",length(seq(-10,10,0.1))),rep("vectors",length(seq(-10,10,0.1))),rep("algebra",length(seq(-10,10,0.1))),rep("analysis",length(seq(-10,10,0.1))),rep("statistics",length(seq(-10,10,0.1)))))
p1 <-ggplot(out) + geom_line(aes(x=delta,y=KL,color = Entry)) + theme_minimal()
p2 <- ggplot(out1) + geom_line(aes(x=delta,y=Jeffreys,color = Entry)) + theme_minimal()
grid.arrange(p1,p2,ncol=2)
```

Misspecifications of the mean of the algebra exam would have the biggest effect on the distribution of the Gaussian BN, since it leads to the biggest distance between the original and the varied network.

### Perturbation of the covariance matrix

Care must be taken when performing perturbations of the covariance matrix, for two reasons: 

 - the perturbed matrix may not be positive semidefinite; 
 
 - the perturbed matrix may not respect the conditional independences of the underlying BN.
 
Suppose we are interested in assessing the effect of varying the covariance between `Statistics` and `Vectors`.


```{r}
gbn$order
gbn$covariance
```

The parameter of interest correspond to the entry (2,5) of the covariance matrix.

A standard perturbed covariance matrix can be constructed with the `covariance_var` function. Suppose we want to increase the covariance between `Statistics` and `Vectors` by an additive factor of 10. 
```{r}
d <- 10
covariance_var(gbn, entry = c(2,5), delta = d)$covariance
```

The above perturbation made the original network structure not valid for the new covariation matrix. In order to ensure that the perturbed covariance is still valid for the underlying network structure, we can use model-preserving methods. These apply multiplicatively and not additively as standard methods, but we apply the same change in the covariance via the perturbation `delta` defined below. We can construct various covariation matrices using the following commands:

```{r}
delta <- (d + gbn$covariance[2,5])/gbn$covariance[2,5]
total_covar_matrix(ci, entry = c(2,5), delta = delta)
partial_covar_matrix(ci, entry = c(2,5), delta = delta)
row_covar_matrix(ci, entry = c(2,5), delta = delta)
col_covar_matrix(ci, entry = c(2,5), delta = delta)
```

Importantly, notice that standard methods are applied to objects of class `gbn`, whilst model-preserving methods operate over `ci` objects.


For any of the four available methods (`total`, `partial`, `row` and `column`) the perturbed covariance matrix can be calculated with the function `model_pres_cov`. For instance in the case of a partial covariation:
```{r}
model_pres_cov(ci, type = "partial", entry = c(2,5), delta = delta)$covariance
```

Having constructed various covariation matrices, we can assess how far apart the original and the perturbed distributions are for various covariation methods. Available dissimilarity measures are Frobenius norm (`Fro`), Kullback-Leibler divergence (`KL`) and Jeffrey's divergence (`Jeffreys`). Let's consider the Frobenius norm.
```{r out.width = "50%"}
d <- seq(-10, 10, 0.1)
delta <- (d + gbn$covariance[2,5])/gbn$covariance[2,5]
cov_stand <- Fro(gbn, entry = c(2,5), delta = d)
cov_col <- Fro(ci, type = "column", entry = c(2,5), delta = delta)
grid.arrange(plot(cov_stand), plot(cov_col), ncol = 2)
```

As for the mean, we can check which entry of the covariance matrix has the biggest impact if varied. For simplicity here we pick the standard method only (code not shown).

```{r,echo=FALSE, out.width="50%"}
d <- seq(-5,5,0.1)
standard11 <- Jeffreys(gbn,"covariance", c(1,1), d)$Jeffreys
standard12 <- Jeffreys(gbn,"covariance", c(1,2), d)$Jeffreys
standard13 <- Jeffreys(gbn,"covariance", c(1,3), d)$Jeffreys
standard14 <- Jeffreys(gbn,"covariance", c(1,4), d)$Jeffreys
standard15 <- Jeffreys(gbn,"covariance", c(1,5), d)$Jeffreys
standard22 <- Jeffreys(gbn,"covariance", c(2,2), d)$Jeffreys
standard23 <- Jeffreys(gbn,"covariance", c(2,3), d)$Jeffreys
standard24 <- Jeffreys(gbn,"covariance", c(2,4), d)$Jeffreys
standard25 <- Jeffreys(gbn,"covariance", c(2,5), d)$Jeffreys
standard33 <- Jeffreys(gbn,"covariance", c(3,3), d)$Jeffreys
standard34 <- Jeffreys(gbn,"covariance", c(3,4), d)$Jeffreys
standard35 <- Jeffreys(gbn,"covariance", c(3,5), d)$Jeffreys
standard44 <- Jeffreys(gbn,"covariance", c(4,4), d)$Jeffreys
standard45 <- Jeffreys(gbn,"covariance", c(4,5), d)$Jeffreys
standard55 <- Jeffreys(gbn,"covariance", c(5,5), d)$Jeffreys
dtot <- rep(d,15)
l <- length(d)
entry <- c(rep("mechanics",l),rep("mechanics/vectors",l),rep("mechanics/algebra",l),rep("mechanics/analysis",l),rep("mechanics/statistics",l),rep("vectors",l),rep("vectors/algebra",l),rep("vectors/analysis",l),rep("vectors/statistics",l), rep("algebra",l), rep("algebra/analysis",l),rep("algebra/statistics",l),rep("analysis",l),rep("analysis/statistics",l),rep("statistics",l))
Jef <- c(standard11$Jeffreys,standard12$Jeffreys,standard13$Jeffreys,standard14$Jeffreys,standard15$Jeffreys,standard22$Jeffreys,standard23$Jeffreys,standard24$Jeffreys,standard25$Jeffreys,standard33$Jeffreys,standard34$Jeffreys,standard35$Jeffreys,standard44$Jeffreys,standard45$Jeffreys,standard55$Jeffreys)
out <- data.frame(d = dtot, entry = entry, Jeffreys = Jef)
ggplot(out) + geom_line(aes(x = d, y = Jeffreys, color=entry))
```

From the above plot we can notice that the less robust entries of the covariance matrix are the variance of `algebra`, the covariance between `algebra` and `analysis`, and the covariance between `algebra` and `vectors`.

Another method to quickly have an overview of the effect of all parameters is `KL_bounds` which creates a table with upper bounds to the Kullback-Leibler divergence for all entries of the covariance matrix and all covariation methods.

```{r}
KL_bounds(ci, delta = 1.2)
```

By looking at the standard method column, we have the confirmation that the 11th entry, corresponding to `algebra/analysis`, is the most critical for the robustness of the network.





