---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bnmonitor

`bnmonitor` is a package for sensitivity analysis and robustness in Bayesian networks (BNs).

## Installation 

The package `bnmonitor` can be installed from CRAN using the command 
```{r warning = FALSE, message = FALSE, eval = F}
install.packages("bnmonitor")
```
and loaded in R with
```{r}
library(bnmonitor)
```
Note that `bnmonitor` requires the package `gRain` which, while on CRAN, depends on packages that are on Bioconductor both directly and through the `gRbase` package, which depends on `RBGL`:
```{r eval =F}
install.packages("BiocManager")
BiocManager::install(c("graph", "Rgraphviz", "RBGL"))
install.packages("gRain")
```

## Overview

`bnmonitor` provides a suite of function to investigate either a data-learnt or an expert elicited BN. Its functions can be classified into three main areas:

 - *Robustness in discrete BNs*: checking how well a BN represents a dataset;
 - *Sensitivity in discrete BNs*: assessing the effect of changes in the discrete BN's probabilities;
 - *Sensitivity in continuous BNs*: assessing the effect of changes in the continuous BN's probabilities, either in the standard or model-preserving framework

<!-- The prequential diagnostics examine the forecasts that flow from a model in sequence. -->
<!-- Each monitor given below indicates the probability of a particular observation based on the previous observations and the model structure.  -->
<!-- In the prequential mindset, we compute a probability of each subsequent observation based on all previous data points.  -->
<!-- These observations are then scored, and in this package we use the logarithmic score function. -->
<!-- The observations are then standardized to give a z-score statistic.  -->
<!-- Following the recommendation of Cowell (2007), scores indicate a poor fit where |z| > 1.96  -->

<!-- We demonstrate the efficacy of the prequential monitors with the Asia data set from the bnlearn package. Details of the variables (nodes) can be found in the documentation for bnlearn. -->
