---
title: "Robustness to Data"
output: rmarkdown::html_vignette
description: >
  Learn how bnmonitor constructs monitors for robustness to data
vignette: >
  %\VignetteIndexEntry{Robustness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = 'center'
)
```


Consider the `asia` dataset available from `bnlearn`. Details of the variables can be found in the `bnlearn` documentation.

``` {r}
library(bnlearn)
data(asia)
summary(asia)
```

Consider two candidate BN models (Lauritzen and Spiegelhalter, 1988), which only differ in the edge from `S` to `D`.
``` {r out.width="50%"}
asia_bn <- bnlearn::model2network("[A][S][T|A][L|S][B|S][D|B:E][E|T:L][X|E]")
bnlearn::graphviz.plot(asia_bn) 
asia_bn_alt <-  bnlearn::model2network("[A][S][T|A][L|S][B|S][E|T:L][X|E][D|B:E:S]")
bnlearn::graphviz.plot(asia_bn_alt) 
```

### Global monitor

<!-- The global monitor is equivalent to the Bayes Factor and offers a good assessment of the model as a whole.  -->
<!-- It is primarily useful for differentiating between candidate models as the global monitor pinpoints the nodes with different contributions for two candidate models. -->
<!-- We can assess the log likelihood contribution from each of the nodes and determine which nodes contribute the most to the global monitor. -->
<!-- global.monitor.graph provides a quick visual for the entire network.  -->
<!-- The darker the color, the more substantial the contribution of that node to the log likelihood. -->

<!-- The global monitor assesses the probability of observing the data given the model and as such does not particularly fit the prequential view.  -->
<!-- However, it is a quick and useful first diagnostic assessment. -->

A first useful diagnostic is the `global_monitor`, reporting the contribution of each vertex to the log-likelihood of the model.

``` {r}
library(bnmonitor)
asia <- asia[sample(1:nrow(asia),50),]
glob_asia <- node_monitor(dag = asia_bn, df = asia, alpha = 3)
glob_asia_alt <- node_monitor(dag = asia_bn_alt, df = asia, alpha = 3)
glob_asia
glob_asia_alt
```

In the alternative model, Dysnopea contributes more to the log-likelihood. 

Global monitors can be plotted giving a quick view of the decomposition of the log-likelihood. The darker the color, the more substantial the contribution of a vertex.

```{r }
plot(glob_asia)
```

### Node monitor 

There are two variants of node monitors.

  - The marginal node monitor computes the probability of the $i$th observation in the data set in turn after passing the evidence of the $i-1$th cases in the data set.
  
  - The conditional node monitor computes the probability of the $i$th observation in the data set after passing evidence of the $i-1$th cases in the data set, and the evidence for all nodes in the $i$th case except the node of interest.

As a quick survey of the nodes, the `node_monitor` command computes the marginal and conditional monitors for the final observation in the data set. 

``` {r echo=TRUE, message = FALSE, warning = FALSE}
node_asia <- final_node_monitor(dag = asia_bn, df = asia)
node_asia
```

The scores indicate a poor fit of the probability distributions specified for the Smoking, Bronchitis, and Dysnopea nodes, since these are larger than 1.96 in absolute value. Plots can also be created to give a visual counterpart of the node monitors.

```{r }
plot(node_asia, which = "marginal")
```

<!-- The modeller may decide which of these distributions is of the most interest.  -->
<!-- For the purposes of this vignette, we assume that we would like to check the forecasts for Dysnopea.  -->

As an illustration we investigate further the fit of the variable Dysnopea.

The sequential marginal monitor `seq_marg_monitor` gives us a closer look at which particular forecasts in the data set might cause this poor fit. We examine the sequential monitor for both candidate models.


``` {r warning=FALSE}
seq_asia <- seq_marg_monitor(dag = asia_bn, df = asia, node.name = "D")
seq_asia_alt <- seq_marg_monitor(dag = asia_bn_alt, df = asia, node.name = "D")
seq_asia
seq_asia_alt
```


```{r warning=FALSE}
plot(seq_asia)
plot(seq_asia_alt)
```

Both monitors indicate that for some observations there is a poor fit (score above 1.96 in absolute value). In particular for the alternative models the marginal monitor has larger values in absolute value. 

A similar analysis can be conducted with `seq_marg_monitor`, which would show that the model fits well (not reported here).


### Parent Child monitor 

Once a vertex has been identified as a poor fit, further investigation can be carried out to check for which values of the parents the model provides a bad representation. This can be achieved with the `seq_pa_ch_monitor` function.

As an illustration consider the `asia_bn` BN, the vertex `D` (Dysnopea), the parent variable `B` (Bronchitis) which can take values `yes` and `no`.

``` {r warning=FALSE, out.width="50%"}
asia_pa_ch1 <- seq_pa_ch_monitor(dag = asia_bn, df = asia, node.name = "D", pa.names =  "B", pa.val =  "yes", alpha = 3)
asia_pa_ch1
asia_pa_ch2 <- seq_pa_ch_monitor(dag = asia_bn, df = asia, node.name = "D", pa.names = "B", pa.val = "no", alpha = 3)
asia_pa_ch2
plot(asia_pa_ch1)
plot(asia_pa_ch2)
```

For this model, Dysnopea is adequately modeled for both values of Bronchitis, since most scores largely fall in the recommended range.

### Influential observations

The last robustness tool is the absolute value of the log-likelihood ratio between a model learnt without one observation and the one learnt with the full dataset. Larger values are associated to atomic events which influence the structural learning. 


```{r warning=FALSE, out.width="50%"}
influence <- influential_obs(dag = asia_bn, df = asia, alpha = 3)
head(influence)
plot(influence)
```
