---
title: "Sensitivity Analysis in Discrete BNs"
output: rmarkdown::html_vignette
description: >
  Learn how bnmonitor implements sensitivity analysis for discrete BNs
vignette: >
  %\VignetteIndexEntry{Sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = 'center'
)
```


The functionalities of `bnmonitor` for sensitivity analysis in Discrete Bayesian Networks are illustrated using the `fire_alarm` network bundled within the package representing a simple alarm system for fire accidents. 

```{r }
library(bnlearn)
library(bnmonitor)
graphviz.plot(fire_alarm)
```

The network consists of 6 nodes (`Fire`, `Smoke`, `Tampering`, `Alarm`, `Leaving` and `Report`) of either two or three levels. 

### Covariation schemes

The aim of a sensitivity analysis is to assess how parameter changes affect output probabilities of interest. When one parameter changes than others need to be adjusted and this can be done in many ways, including  `proportional`, `uniform` and `order-preserving` schemes. Suppose we vary the probability that `Alarm = FALSE` given that `Fire = TRUE` and `Tampering = FALSE` from 0.01 to 0.1. The resulting probabilities from the proportional covariation scheme can be computed as follows.

```{r}
proportional_covar(bnfit = fire_alarm, node = "Alarm", value_node = "FALSE", value_parents = c("TRUE","FALSE"), new_value = 0.1)$Alarm$prob
```

The functions `uniform_covar` and `orderp_covar` have the same inputs and return the BN where probabilities are updated using uniform and order-preserving covariation, respectively.

### Sensitivity functions

Suppose we are interested in the probability that `Report = TRUE` and how this varies with respect to the probability that `Alarm = FALSE` conditional on `Fire = TRUE` and `Tampering = FALSE`. This can be computed with the function `sensitivity`.

```{r  warning = FALSE}
library(ggplot2)
sens <- sensitivity(bnfit = fire_alarm, interest_node = "Report", interest_node_value =  "TRUE" , node = "Alarm" , value_node = "FALSE", value_parents = c("TRUE","FALSE"), new_value = "all", covariation = "all")
plot(sens) + theme(legend.title = element_text(size = 5), legend.text = element_text(size = 4), legend.position = "bottom")
```

For all covariation schemes as the conditional probability of `Alarm = FALSE` increases, the probability of `Report = TRUE` decreases. The uniform covariation scheme is reported in red, the proportional in green and the order-preserving in blue.

The function sensitivity also allows for conditional probabilities of interest, not only marginals, which can be set by the options `evidence_nodes` and `evidence_states`.

For such changes in the probability distribution of a BN, we can compute the distance between the original and the varied networks using the Chan-Darwiche distance (`CD`) and the Kullback-Leibler divergence (`KL`). As an example consider `Alarm = MALFUNCTIOn` conditional on `Fire = TRUE` and `Tampering = FALSE`.

```{r warning = FALSE}
cd_distance <- CD(fire_alarm, node = "Alarm", value_node = "MALFUNCTION", value_parents = c("TRUE","FALSE"), new_value = "all", covariation = "all") 
kl_divergence <- KL(fire_alarm, node = "Alarm", value_node = "MALFUNCTION", value_parents = c("TRUE","FALSE"), new_value = "all", covariation = "all") 
plot(cd_distance) + theme(legend.title = element_text(size = 5), legend.text = element_text(size = 4), legend.position = "bottom")
plot(kl_divergence) + theme(legend.title = element_text(size = 5), legend.text = element_text(size = 4), legend.position = "bottom")
```

As expected proportional covariation has the smallest distance/divergence of all covariation schemes.

### Sensitivity queries

Another task in sensitivity analysis is to identify parameter changes that meet a specific constraint. As an example suppose we want the probability `Report = TRUE` to be equal to 0.4. Which parameters could we change to obtain such a probability? The answer can be obtained using the function `sensquery`.

```{r}
sensquery(bnfit = fire_alarm, interest_node = "Report", interest_node_value = "TRUE", new_value = 0.4)
```

There are five parameter changes that would lead to a probability of `Report = TRUE` of 0.4. These are ordered by the value of the CD distance between the original and the new network, from smallest to largest.
